fairmetrics: An R package for group fairness evaluation

### 2025-08-21 utils/joss_paper/paper.Rmd
updating get_fairness_metrics to return a dataframe

---
title: 'fairmetrics: An R package for group fairness evaluation'
tags:
  - R
  - Fairness
  - Machine Learning
  - Software
authors:
  - name: Benjamin Smith
    orcid: 0009-0007-2206-0177
    affiliation: 1
  - name: Jianhui Gao
    orcid: 0000-0003-0915-1473
    affiliation: 1
  - name: Benson Chou
    orcid: 0009-0007-0265-033X
    affiliation: 1
  - name: Jessica Gronsbell
    orcid: 0000-0002-5360-5869
    affiliation: 1
affiliations:
  - name: "Department of Statistical Science, University of Toronto"
    index: 1
date: "2025-06-10"
bibliography: paper.bib
output:
  # rticles::joss_article:
  #   keep_tex: true
  md_document:
     #keep_tex: true
     preserve_yaml: TRUE
     variant: "markdown_strict"
journal: JOSS
---

@@

$performance
                                    Metric GroupFemale GroupMale
1                 Positive Prediction Rate        0.17      0.08
2                      False Negative Rate        0.38      0.62
3                      False Positive Rate        0.08      0.03
4            Avg. Predicted Positive Prob.        0.46      0.37
5            Avg. Predicted Negative Prob.        0.15      0.10
6                Positive Predictive Value        0.62      0.66
7                Negative Predictive Value        0.92      0.90
8                              Brier Score        0.09      0.08
9                                Accuracy        0.87      0.88
10 (False Negative)/(False Positive) Ratio        1.03      3.24

$fairness
                           Metric Difference    95% Diff CI Ratio 95% Ratio CI
1              Statistical Parity       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
2               Equal Opportunity      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
3             Predictive Equality       0.05   [0.02, 0.08]  2.67 [1.38, 5.15]
4      Balance for Positive Class       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
5      Balance for Negative Class       0.05   [0.03, 0.07]  1.50 [1.29, 1.75]
6      Positive Predictive Parity      -0.04  [-0.21, 0.13]  0.94 [0.71, 1.24]
7      Negative Predictive Parity       0.02  [-0.02, 0.06]  1.02 [0.98, 1.07]
8              Brier Score Parity       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
9         Overall Accuracy Parity      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
10             Treatment Equality      -2.21  [-4.44, 0.02]  0.32  [0.14, 0.7]
```

>>>>

---
title: 'fairmetrics: An R package for group fairness evaluation'
tags:
  - R
  - Fairness
  - Machine Learning
  - Software
authors:
  - name: Benjamin Smith
    orcid: 0009-0007-2206-0177
    affiliation: 1
  - name: Jianhui Gao
    orcid: 0000-0003-0915-1473
    affiliation: 1
  - name: Benson Chou
    orcid: 0009-0007-0265-033X
    affiliation: 1
  - name: Jessica Gronsbell
    orcid: 0000-0002-5360-5869
    affiliation: 1
affiliations:
  - name: "Department of Statistical Science, University of Toronto"
    index: 1
date: "2025-06-10"
bibliography: paper.bib
output:
  rticles::joss_article:
    keep_tex: true
  md_document:
     #keep_tex: true
     preserve_yaml: TRUE
     variant: "markdown_strict"
journal: JOSS
---

@@

           Fairness Assesment                                  Metric
1          Statistical Parity                Positive Prediction Rate
2           Equal Opportunity                     False Negative Rate
3         Predictive Equality                     False Positive Rate
4  Balance for Positive Class           Avg. Predicted Positive Prob.
5  Balance for Negative Class           Avg. Predicted Negative Prob.
6  Positive Predictive Parity               Positive Predictive Value
7  Negative Predictive Parity               Negative Predictive Value
8          Brier Score Parity                             Brier Score
9     Overall Accuracy Parity                                Accuracy
10         Treatment Equality (False Negative)/(False Positive) Ratio

   GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
1         0.17      0.08       0.09   [0.05, 0.13]  2.12 [1.49, 3.04]
2         0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
3         0.08      0.03       0.05   [0.02, 0.08]  2.67  [1.4, 5.08]
4         0.46      0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.42]
5         0.15      0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.74]
6         0.62      0.66      -0.04  [-0.21, 0.13]  0.94 [0.72, 1.22]
7         0.92      0.90       0.02  [-0.02, 0.06]  1.02 [0.98, 1.07]
8         0.09      0.08       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
9         0.87      0.88      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
10        1.03      3.24      -2.21 [-4.38, -0.04]  0.32 [0.15, 0.68]
```
